Live Functional Programming with Typed Holes

### 2018-07-07 examples-1.tex
fix spelling typo
submission
small edits to 2.1

Consider the perspective of a teacher in the midst of developing a \Hazel{} notebook
to compute final student grades at the end of a course.
The top of \autoref{fig:grades-example} depicts the cell containing the incomplete program that 
the teacher has written so far (we omit irrelevant parts of the UI). 
%

At the top of this program, the teacher defines a
record type, \lismall{Student}, for recording a student's course data---here,
the student's name, of type
\lismall{string}, and, for simplicity, three grades, each of type \lismall{float}. The variable \lismall{students} binds a list of such student records. For simplicity, we include only three example students.
At the bottom of the program, the teacher maps a function \li{weighted_average} over this student data (\li{map} is the standard map function over lists, not shown), intending to compute a final weighted average for each student.
However, the program is incomplete because the teacher has not yet completed the body of the \lismall{weighted_average} function. This pattern is well-established: programmers often consume a function before implementing it.

Thusfar in Fig.~\ref{fig:grades-example}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator. The cursor now sits at an \emph{empty hole}, as indicated by the vertical bar and the green background. Each hole has a unique name (generated automatically in \Hazel), here simply \li{1}. 
%In this case, there is only one hole in the program, but in general, there may be many such holes.

Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer.
In response, the programmer might insert a temporary universal placeholder, e.g. by raising an exception as suggested in Sec.~\ref{sec:intro} (note however that Elm does not support exceptions, by design). This would cause typechecking to succeed. 
However, when the programming environment attempts to evaluate the program, 
evaluation would proceed only as far as the first \li{map} iteration, 
which would call into \li{weighted_average} and then fail when attempting to evaluate the hole.%
\footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.

\Hazel{} does not take this ``exceptional'' interpretation of holes. Instead, evaluation continues past the hole, treating it as an opaque expression of the appropriate type. The result, shown at the bottom of Fig.~\ref{fig:grades-cell-mockup}, is a list of length $3$, confirming that \li{map} does indeed behave as expected in this regard despite the teacher having provided an incomplete argument. Furthermore, each element of the resulting list has been evaluated as far as possible, i.e. the arithmetic expression \li{30.0*hw} has been evaluated for each corresponding value of \li{hw}, as expected. Evaluation cannot proceed any further because holes appear as addends. We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a hole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in critical positions.

Given this indeterminate result, the teacher might take notice of the magnitude of the numbers being computed, e.g. \li{2640.0} and \li{2280.0}, and
decide immediately that a mistake has been made: the teacher wants to compute a 
weighted average between \li{0.0} and \li{100.0}, and so the correct
constant is \li{0.30} rather than \li{30.0}. (There are of course other possible fixes.)

Although these observations might 
save only a small amount of time in this case, 
% relative to the 
% situation where the teacher notices this mistake only once  
% the \li{weighted_average} function has been completed, 
it demonstrates
the broader motivations of live programming: continuous feedback about the dynamic behavior of the 
program can help confirm the mental model 
that the programmer has developed (in this case, regarding the behavior of \li{map}), or else help quickly dispel 
inconsistencies between the programmer's expectations and the 
actual behavior of the program (in this case, the arithmetic expression).

% The result itself is not the only live feedback that \Hazel reports to the programmer.
% Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
\Hazel also helps programmers reason about the dynamic behavior of bound variables in scope at a hole via the \emph{live context inspector}, normally displays it as a sidebar but shown disembodied in three states in Fig.~\ref{fig:grades-sidebar}. 
In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments.

New to our approach are the values associated with each binding.
These values come from the environments associated with each hole instance in the result. In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}. An environment is a partial mapping of each variable in scope to a value. We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical foundations detailed in Sec.~\ref{sec:calculus}). The programmer can select different closures by clicking on a hole instance (here, instance \li{1:1} was selected by default, as indicated by the purple outline), or cycle rapidly through all of the closures for the hole at the cursor by clicking the arrows, e.g. $\blacktriangleright$, at the bottom of the context inspector, or corresponding hotkeys.

\vspace{-4px}
\subsection{Example 2: Recursive Functions}\label{sec:qsort1}
\vspace{-2px}

@@

The programmer has called \li{qsort} on an example list. However, 
the indeterminate result in this case is rather uninteresting: it simply confirms that for this example, evaluation went through the recursive case of \li{qsort}, which ends with hole \li{1}. The live context inspector, on the other hand, provides a wealth of feedback about the dynamic behavior of this incomplete program. For example, the programmer can confirm that
the lists \li{smaller} and \li{bigger} are appropriate named.

>>>>

Consider the perspective of a teacher in the midst of developing a \Hazel{} notebook
to compute final student grades at the end of a course.
\autoref{fig:grades-cell-mockup} depicts the cell containing the incomplete program that 
the teacher has written so far (we omit irrelevant parts of the UI). 
%

At the top of this program, the teacher defines a
record type, \lismall{Student}, for recording a student's course data---here,
the student's name, of type
\lismall{string}, and, for simplicity, three grades, each of type \lismall{float}. Next, the teacher constructs a list of student records, binding it to the variable \lismall{students}. For simplicity, we include only three example students.
At the bottom of the program, the teacher maps a function \li{weighted_average} over this student data (\li{map} is the standard map function over lists, not shown), intending to compute a final weighted average for each student.
However, the program is incomplete because the teacher has not yet completed the body of the \lismall{weighted_average} function. This pattern is quite common: programmers often consume a function before implementing it.

Thusfar in the body of \lismall{weighted_average}, the teacher has decomposed the function argument into variables by record destructuring, then multiplied the homework grade, \li{hw}, by \li{30.0} and finally inserted the \li{+} operator. The cursor now sits at an \emph{empty hole}, as indicated by the vertical bar and the green background. Each hole has a unique name (generated automatically in \Hazel), here simply \li{1}. 
%In this case, there is only one hole in the program, but in general, there may be many such holes.

Let us briefly digress: in a conventional ``batch'' programming system, writing \li{30.0*hw +} by itself would simply cause parsing to fail and there would be no static or dynamic feedback available to the programmer at all.
In response, the programmer might insert a temporary universal placeholder, e.g. by raising an exception as suggested in Sec.~\ref{sec:intro} (note however that Elm does not support exceptions, by design). This would cause typechecking to succeed. 
However, when the programming environment attempts to evaluate the program, 
evaluation would proceed only as far as the first \li{map} iteration, 
which would call into \li{weighted_average} and then fail when attempting to evaluate the hole.%
\footnote{In a lazy language, like Haskell, the result would be much the same because the environment forces the result for printing.}
While the stack trace would confirm that \li{weighted_average} has been called, this is not entirely satisfying.

\Hazel{} does not take this ``exceptional'' interpretation of holes. Instead, evaluation continues past the hole, treating it as an opaque expression of the appropriate type. The result, shown at the bottom of Fig.~\ref{fig:grades-cell-mockup}, is a list of length $3$, confirming that \li{map} does indeed behave as expected in this regard despite the teacher having provided an incomplete argument. Furthermore, each element of the resulting list has been evaluated as far as possible, i.e. the arithmetic expression \li{30.0*hw} has been evaluated for each corresponding value of \li{hw}, as expected. Evaluation cannot proceed any further because holes appear as addends. We say that each of these addition expressions is an \emph{indeterminate} sub-expression, and the result as a hole is therefore also indeterminate, because it is not yet a value, nor can it take a step due to holes in elimination positions.

At this point, the teacher might take notice of the magnitude of the numbers being computed, e.g. \li{2640.0} and \li{2280.0}, and
realize immediately that a mistake has been made: the teacher wants to compute a 
weighted average between \li{0.0} and \li{100.0}, and so the correct
constant is \li{0.30} rather than \li{30.0}. (There are of course other possible fixes.)

Although these observations might 
save only a small amount of time in this case, 
% relative to the 
% situation where the teacher notices this mistake only once  
% the \li{weighted_average} function has been completed, 
it demonstrates
the broader motivations of live programming: continuous feedback about the dynamic behavior of the 
program can help confirm the mental model 
that the programmer has developed (in this case, regarding the behavior of \li{map}), and also help quickly dispel 
inconsistencies between the programmer's expectations and the 
actual behavior of the program (in this case, the magnitude of the arithmetic expressions being computed).

% The result itself is not the only live feedback that \Hazel reports to the programmer.
% Programmers also keep track of the names and types of the variables that are in scope at the cursor. 
Going further, \Hazel helps programmers reason about the dynamic behavior of bound variables in scope at a hole via the \emph{live context inspector}, normally displayed  as a sidebar but shown disembodied in three different states in Fig.~\ref{fig:grades-sidebar}. 
In all three states, the live context inspector displays the names and types of the variables that are in scope, as is a common feature of programming environments.

New to our approach are the values associated with each binding.
These values come from the environments associated with each hole instance in the result. In this case, there are three instances of hole \li{1} in the result, numbered sequentially: \li{1:1}, \li{1:2} and \li{1:3}, arising from the three calls to \li{weighted_averages}. An environment is a partial mapping of each variable in scope to a value. We call a hole instance paired with an  environment a \emph{hole closure}, by analogy to function closures (and also due to the logical foundations detailed in Sec.~\ref{sec:calculus}). The programmer can select different closures by clicking on a hole instance in the result (the first instance, \li{1:1}, is selected by default, as indicated by the solid purple outline), or cycle rapidly through all of the closures for the hole at the cursor by clicking the arrows, e.g. $\blacktriangleright$, at the bottom of the context inspector, or by using hotkeys.

\vspace{-4px}
\subsection{Example 2: Recursive Functions}\label{sec:qsort1}\label{sec:paths}
\vspace{-2px}

@@

The programmer has called \li{qsort} on an example list. However, 
the indeterminate result in this case is rather uninteresting: it simply confirms that for this example, evaluation went through the recursive case of \li{qsort}, which ends with hole \li{1}. The live context inspector, on the other hand, provides a wealth of feedback about the dynamic behavior of this incomplete program. For example, the programmer can confirm that
the lists \li{smaller} and \li{bigger} are appropriately named.

### 2018-07-07 examples.tex
small edits to 2.1

This section gives an example-driven overview of our  approach as implemented in  \Hazel, a browser-based lab notebook environment being developed by \cite{HazelnutSNAPL}. The \Hazel user interface is  based roughly on IPython/Jupyter \cite{PER-GRA:2007} and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}), a popular pure functional programming language similar to ``core ML'' \cite{czaplicki2012elm,Elm}, with which we assume basic  familiarity.
For the sake of 
exposition, we have ``post-processed'' the screenshots in this section to make use of a few language conveniences from \Elm~that were not available in \Hazel as of this writing, e.g. pattern matching in function arguments, list  notation, and record labels (currently there are only tuples). These conveniences are entirely orthogonal to the contributions of this paper: all of the live programming features demonstrated in this section have been implemented, both in \Hazel itself and in the simplified companion implementation of the core calculus in the supplement (see Sec.~\ref{sec:implementation}).

>>>>

This section gives an example-driven overview of our  approach as implemented in  \Hazel, a browser-based lab notebook environment being developed by \citet{HazelnutSNAPL}. The \Hazel user interface is  based roughly on IPython/Jupyter \cite{PER-GRA:2007} and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume basic  familiarity.
For the sake of 
exposition, we have ``post-processed'' the screenshots in this section to make use of a few language conveniences from \Elm~that were not available in \Hazel as of this writing, e.g. pattern matching in function arguments, list  notation, and record labels (currently there are only tuples). These conveniences are entirely orthogonal to the contributions of this paper: all of the live programming features shown in this section have been implemented, both in \Hazel itself and in the simplified companion implementation of the core calculus included in the supplement (see Sec.~\ref{sec:implementation}).
